---
title: "Class Reference"
author: "Derek Willis"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
print("Hello puppy!")
```

## Introduction

Consider this a personal guide to the commands and functions you will learn. In general, when you come across an R command or function that you want to remember, put it in here along with a description of what it does and when you'd use it.

### How to set the working directory

The command to set a working directory is setwd(). For example:

```{r}
setwd("~/code/data_journalism_2023_spring")
```

### How to install and load the tidyverse

```{r}
#install.packages("tidyverse")
library(tidyverse)
```

### How to read in a CSV file from the Internet

```{r}
wes_moore_contributions <- read_csv("https://raw.githubusercontent.com/stephenneukam/CNS_Annapolis/main/Campaign_finance/Moore_ContributionsList.csv")
```

### How to count rows

You use the `n()` function to count rows, like so:

```{r}
dataframe |>
  group_by(some_column) |>
  summarize(count = n())
```

How to export FROM R Studio 

write_csv

How to mutate a date from <chr> to <date>  
```{r}

md_expenses <- md_expenses |> mutate(expenditure_date = mdy(expenditure_date))


```

remove all entries where column (payee_name in this case) has NA values 

```{r}
md_expenses |> filter(!is.na(payee_name))
```

if you DID want to see all the ones where the payee is NA, you'd remove the ! 
```{r}
md_expenses |> filter(is.na(payee_name))

```

or use drop_na()
```{r}
dataset |> 
  drop_na(payee_name)
```

NA is not LITERALLY what the value is, which is why you can't str_detect(payee_name = "NA"). NA is a special designation by R, a lack of value in the cell. It is greyed out in the dataset for this reason. 

ALWAYS TELL CHAT GPT THAT YOU ARE WORKING IN THE TIDYVERSE! SO it doesn't give you random code from other libraries. 


Extract an hour from a time column and create a new column called hour 
```{r}
earthquakes_hourly <- earthquakes |> 
  mutate(hour = hour(time))
```

Renaming a column nameusing rename ()

```{r}
dataset |> 
  rename(new_column_name = old_column_name)
```

Replace or change the contents of a column, you're using mutate. 

str_replace_all() a literal period is '\\." 

str_squish gets rid of leading or trailing spaces 
```{r}
|> 
  mutate(jurisdiction = str_squish(str_replace_all(jurisdiction,'\\.','')))
```

Combining datasets 
```{r}
county_voters_combined <- bind_rows(list(county_voters_2016, county_voters_2018, county_voters_2020, county_voters_2022))

```

glimpse gives you summaries 
```{r}
glimpse(maryland_zcta)
```

changing datatype from dbl to character **You always want zip codes to be characters, not numbers. 

```{r}
mutate(ZCTA5N = as.character(ZCTA5N))

```

changing datatype from character to dbl 


```{r}
mutate(ZCTA5N = as.numeric(ZCTA5N))

```
To generate a new column called month, I first tried mutate() with month() but that was giving me a number 1-12 to label the month instead of its name. When I typed month I could see that R was suggesting other possible functions for me so I just kept trying until I was successful with months.POSIXt(). That function apparently uses the base library so I loaded that above. 

```{r}
county_911_combined <- county_911_combined |> 
  mutate(month = months.POSIXt(date))
```


renaming a column
```{r}
dataset |> 
  rename (new_column_name = old_column_name)
```

**Task** Run the following code to join Maryland population data and the 2022 voter registration data together using the common column county as the key. How many rows are there now? How many *should* there be? 

**Answer**

```{r}
# Joining 2 dataframes with nrow included to show row total
maryland_population |> left_join(county_voters_2022, join_by("COUNTY")) |> nrow()