---
title: "lab_08"
author: "caley fox shannon"
date: "2023-11-08"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings. 

**Task** Create a codeblock and load appropriate packages and settings for this lab. We'll be making some charts, working with dates and retrieving Census data.

```{r}
library(tidyverse)
library(lubridate)
library(janitor)
library(tidycensus)
```

## Load and modify data

**Task** Create a codeblock and load the following data from the data folder:

* Prince George's County 911 Overdose Calls

You will need to create columns for the date, week and month based on the existing `datetime` column.

```{r}

# Load in data. 
pg_911 <- read_csv("data/prince_georges_2022_overdoses.csv")

# Create columns for the date, week and month. I did not ask Chat GPT for this code, I just messed around guessing what tidyverse might call these functions until I got it right. 

pg_911 <- pg_911 |> 
  mutate(
    date = day(datetime), 
    week = week(datetime), 
    month = month(datetime, label=TRUE))
```

## Questions 

**Q1.** Which month saw the greatest percentage of total calls? Create a dataframe that calculates the percentage of all calls that each month's calls represents. Make a bar chart from that dataframe. Your bar chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable bars - the values shouldn't be overlapping

```{r}
# Create a dataframe that calculates the percentage of all calls that each month's calls represents. 
monthly_pg_911 <- pg_911 |> 
  group_by(month)|> 
  summarise(monthly_calls = n())|> 
  mutate(pct_of_calls = (monthly_calls/ sum(monthly_calls))*100)|> 
  arrange(desc(pct_of_calls))

# Make a bar chart from that dataframe.

monthly_pg_911 |> 
  ggplot() +
  geom_bar(aes(x=month, weight=pct_of_calls)) + 
  labs(
    title="PG County overdoses spike in July, November, December",
    x = "Months, 2022",
    y = "Overdose Calls Per Month",
    caption = "source: Prince George's County EMS")

```

Compare the results here to those from the pre_lab_08 bar chart - are there major differences in the months with the highest and lowest figures? Describe that below. 

**A1.** December 2022 had the highest percentage of overdose calls from PG County that year at 10.9%. This chart appears to differ greatly from the Baltimore County 911 calls bar chart in the pre lab. While the Baltimore chart has far greater calls across all months, the trends do not appear to correspond to overdoses in PG County. Perhaps there is something particular when how overdoses occur as opposed to 911 calls in general. If I had to hazard a guess, I'd attribute the spikes in overdoses during July, November and December to major holidays. 

**Q2.** Let's visualize this data on a weekly basis using a line chart. As in Q1, generate a dataframe with the total number of calls for each week, and then create a line chart to show the distribution of calls over time. Your line chart must have:

* A clear title that states the main idea/finding
* Good labels for the x & y axis and a caption for the source, which is Prince George's County EMS.
* Readable labels

Describe the pattern of results; you may want to look at the data to dig into particular outliers.

**A2.** I'm not particularly satisfied with this data in line graph form. For one, I don't think people think in the number of the weeks in the year. If you gave me some data for week 31 of the year, I'd have to think pretty hard about when that is and do some math to figure it out. I tried to use scale_x_date() to add more ticks but it frankly didn't look better to me. 

Another conundrum of measuring 2022 in weeks is that there is a partial 53rd week, during which 4 overdose calls were placed. Because the week is not complete, it appears that the number of calls was staggeringly low. That causes the line graph to dip dramatically at the end of the year, which is probably misleading. 

While there may be an elegant solution for finessing these issues here, I maintain that people don't really think in weeks so it's not worth pursuing. 


```{r}
# Generate a dataframe with the total number of calls for each week. 

weekly_pg_911 <- pg_911 |>
  group_by(week)|> 
  summarise(weekly_calls = n())

weekly_pg_911

# Create a line chart to show the distribution of calls over time.

weekly_pg_911 |>
  ggplot() + 
  geom_line(aes(x=week, y=weekly_calls))+ 
  labs(
    title="PG County Overdoses Peak in Summer",
    x = "Weeks, 2022",
    y = "Overdose Calls Per Week",
    caption = "source: Prince George's County EMS")


```

**Q3.**  A scatterplot is a type of chart that helps us see relationships between two variables. One variable goes on the x axis, the other on the y axis.  For each row/observation in our data, a scatterplot puts a circle (or a "point") where the two variables intersect on a grid. 

Statisticians use scatterplots to show graphically whether one variable is correlated -- related, in a statistical sense -- with another variable.  A classic example is the [relationship between ice cream sales and temperature](https://www.mathsisfun.com/data/scatter-xy-plots.html). The scatterplot below -- press play to load the image -- shows that relationship, that an increase in temperature is associated with an increase in ice cream sales. When it's 12C, sales are 200 dollars, and when it's hotter, 25C, sales are 600 dollars.

```{r}
knitr::include_graphics("https://www.mathsisfun.com/data/images/scatter-ice-cream1.svg")
```

We're going to use a scatterplot a little differently, to get a visual sense of two key variables: 

Our question is: does the median income in a zip code have any relationship to the number of overdose 911 calls in that zip code?

To answer this question, do the following:

1. Generate a dataframe with the number of 911 calls for each zip code.
2. Get data from the Census Bureau showing median household income for Maryland zip codes.
3. Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.
4. Make a scatterplot showing the total calls and median income. I didn't show you how to do this, so look it up! Googling "ggplot scatterplot" is a good start.
5. Give it an appropriate title, source, and x and y axis titles.
6. Add a label for each point that shows the zip code using geom_text() - see some examples of its use at https://ggplot2.tidyverse.org/reference/geom_text.html#ref-examples. Try to make the names as easy to read as possible by avoiding overlap.
7. In the answer space below, describe what you see and answer the questions posed above. In a general sense, what do you think this means? Feel free to consider the actual raw values: how would you report out the main point(s) of this chart?

**A3.**  It looks to me like there is a relatively weak correlation between the median income of a zip code in PG County and the number of 911 calls placed for overdoses when the median income is less than $100,000 annually. But once the median income is greater than $100,000,  the number of overdose calls generally decreases as income increases. There may be something really interesting there: it could be that until a household is living "very comfortably" let's say, income and overdoses are not correlated. 

```{r}

#1. Generate a dataframe with the number of 911 calls for each zip code.

zip_pg_911 <- pg_911 |> 
  group_by(zipcode)|> 
  summarise(calls_per_zipcode = n()) 

#2. Get data from the Census Bureau showing median household income for Maryland zip codes.

census_variables <- load_variables(2019, "acs5", cache=TRUE)

md_income <- get_acs(geography = "zcta",
              variables = c(median_income = "B19013_001"),
              state = "MD",
              year = 2019)

#3. Join those two dataframes on their zip code columns, starting with the 911 calls dataframe.

zip_income_pg_911 <- zip_pg_911 |> 
  inner_join(md_income, by=c('zipcode' = 'GEOID'))|> 
  rename(median_income = estimate)

#4-6. Make a scatterplot showing the total calls and median income. Give it an appropriate title, source, and x and y axis titles. Add a label for each point that shows the zip code using geom_text. 

zip_income_pg_911|>
  ggplot() + 
  geom_point(aes(y=calls_per_zipcode, x=median_income))+ 
  geom_text(aes(y=calls_per_zipcode+5, x=median_income, label=zipcode), check_overlap = TRUE)+
  labs(
    title="High Income Areas Place Fewer Overdose 911 Calls",
    y = "Calls Per Zipcode",
    x = "Median Income",
    caption = "source: Prince George's County EMS")


```
