---
title: "lab_03"
author: "Derek Willis"
date: "2023-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load libraries and establish settings

```{r}
# Turn off scientific notation
options(scipen=999)

# Load the tidyverse
library(tidyverse)

# Load janitor
library(janitor)
```

## TASK: Load data and clean up the column names

Load the two datasets we worked with in the pre_lab: the Maryland payments to vendors and Maryland grants and loans and, using janitor, change the columns so that they are lower-cased and any spaces are replaced with underscores.

```{r}
# Load in the Maryland state agency payments data, making sure that it correctly sets the datatypes using all of the rows

payments <- read_csv("data/State_of_Maryland_Payments_Data__FY2008_to_FY2024.csv") |> 
  clean_names() 

# Load in the Maryland grants and loan data

md_grants_loans <- read_csv("data/State_of_Maryland_Grant_and_Loan_Data__FY2009_to_FY2022.csv") |> 
  clean_names() 

```

## Answer questions

Q1: You want to know how much money state agencies have paid to the Prince George's County School system, but to do that you'll need to find all of the variations of how the school system might be listed. Write code to find rows where the vendor name contains the county's school system, starting with a broad search and then narrowing it down. When you have found as many as possible, save the result to a new dataframe called `pg_schools`. How many records did you find? Look at & search the records in the dataframe while you construct your code.

A1:

```{r}

pg_schools = payments |>
  filter(str_detect(vendor_name, "PRINCE GEORGE") & str_detect(vendor_name, "SCHOOL"))

pg_schools

```

Q2: Using that `pg_schools` dataframe, write code to calculate the total amount of payments to the Prince George's County School system for each fiscal year. Which fiscal year had the largest total amount? Are there any aspects of the results that stand out to you?

A2: 2021 had the largest total amount of payments to the Prince George's County school system. It is interesting that payments haven't either steadily rose or steadily fallen over time, the cash flow seems a bit scrambled chronologically. 

```{r}

pg_schools |> 
  group_by(fiscal_year) |> 
  summarise(total = sum(amount)) |> 
  arrange(desc(total))


```

Q3. You're interested in examining the vendors paid for capital projects at the University of Maryland. Search the payments data to find those payments, then write code to produce a list of each vendor and the total amount it received. Look carefully at that list - what cleaning will it require to ensure that you have only one row per vendor and zip code? Provide at least three examples of vendor names that are similar and should be combined into one. Describe how you might clean that data so that any nearly identical vendors would look identical - what would you change and how?

A3. There are obviously a lot of duplicates in this vendor list, likely due to spelling and punctuation errors. These vendors appear to be duplicates: A MORTON THOMAS & ASSOC INC & A MORTON THOMAS & ASSOCIATES INC; A&I INC & A&I INC.; AEROSOL MONITORING & ANALYSIS INC & AERSOL MONITORING & ANALYSIS INC. 

Is there a code that removes all characters that aren't a letter? That seems like it would help. I'm not sure how to account for the spelling errors though. I believe get_dupes only finds exactly matched entries. 

```{r}

payments |>
  filter(str_detect(agency_name, "CAPITAL PROJECTS- UNIVERSITY OF MARYLAND")) |> 
  group_by (vendor_name) |> 
  summarise (total = sum(amount))

```

Q4: You want to understand how much state grant money went to stem cell research projects each fiscal year. Search the data to find grants that fit that description. Once you find which columns contain the information you want, your code should standardize the data needed to find just the grants for stem cell research in a single filter. Then create a total amount of money for each fiscal year. Write a sentence summarizing how stem cell research funding has changed over time.

A4: The amount that the state of Maryland granted to support stem cell research has steadily declined over time. From 2009 to 2018, the funding was cut by more than half, from over $14.6 million to approximately $6.2 million.  

```{r}

md_grants_loans |>
  filter(str_detect(grantee, "Stem Cell") |str_detect(grantor, "Stem Cell")) |>
  group_by(fiscal_year) |> 
  summarise(total = sum(amount)) |> 
  arrange (desc(total))


```

Q5: Using the Maryland grants and loan data, find records where the zip code is the one where the University of Maryland-College Park is located in and produce a list of grantees from that zip code, along with the total amount they received. Write a paragraph explaining how many grantees are there, what you would need to do to clean the results so that you had one record per grantee and which grantees are most interesting to you?

A5: Eighteen grantees in the 20742 zip code received grants from the state of Maryland. The top grantee is the University of Maryland, but there appear to be grantees that are subsidiaries/offices of the university and should be re-named to be grouped with the university. Capitalization is also an issue. First I would change the grantees to all be displayed in uppercase text. Then I would run code that renames any grantee that contains "University of Maryland," "Unviersity of MD," or "UMD" to be called "University of Maryland." 

```{r}

md_grants_loans |>
  filter(str_detect(zip_code, "20742")) |> 
  group_by(grantee) |> 
  summarise(total = sum(amount)) |> 
  arrange(desc(total))
  
  
```

