---
title: "Grad Project: Violent Crime in Maryland Since 1975 By County"
author: "Caley Fox Shannon"
date: "12/04/23"
output:
  html_document:
    theme: cerulean
    highlight: pygments
    toc: true
    toc_float:
      collapsed: true
      smooth_scroll: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction 

In this notebook, I will analyse the dataset "Violent Crime & Property Crime by County: 1975 to Present" (https://opendata.maryland.gov/Public-Safety/Violent-Crime-Property-Crime-by-County-1975-to-Pre/jwfa-fdxs). 

## Load libraries

```{r echo=FALSE, message=FALSE}

# Turn off scientific notation. 

options(scipen=999)

# Load libraries. 

library(tidyverse)
library(lubridate)
library(janitor)
library(tidycensus)
library(sf)
library(tigris)
```

## Load and Cleaning Data

```{r}

# Load required data.

crime_by_county <- read_csv("data/crime_by_county.csv")

# Clean required data and prepare for analysis if needed. 

crime_by_county <- clean_names(crime_by_county)

crime_by_county <- crime_by_county |> 
  rename(breaking_and_entering = b_e)|> 
  rename(motor_vehicle_theft = m_v_theft)|> 
  rename(breaking_and_entering_per_100_000_people = b_e_per_100_000_people)|> 
  rename(motor_vehicle_theft_per_100_000_people = m_v_theft_per_100_000_people)|> 
  rename(breaking_and_entering_rate_percent_change_per_100_000_people = b_e_rate_percent_change_per_100_000_people) |> 
  rename(motor_vehicle_theft_rate_percent_change_per_100_000_people = m_v_theft_rate_percent_change_per_100_000_people) |> 
  rename(county = jurisdiction)|> 
  mutate(county = str_to_upper(county))

# Find NA values 

sum(is.na(crime_by_county))

```

## Description of Data 

This dataset was compiled from Maryland State Police's annual Uniform Crime Reports. There are 1,104 rows, one per county per year from 1975 to 2020. Based on the number of rows, this dataset is complete and not missing any years for any counties (24 counties * 46 years = 1,104). 

The dataset includes 312 NA values. These NA values are for each county's 1975 row, in the 13 columns that include a change over time (24 counties * 13 columns = 312). This makes sense because 1975 is the first year in the dataset. 

There are 38 columns. Columns include the county, year, population, and 7 major types of crime (murder, rape, robbery, aggravated assault, breaking and entering, larceny theft and motor vehicle theft) and a grand total of crimes. There are also columns that total violent crimes and property theft crimes, as well as calculating per capita rates and percent change per year for all categories. 

## Confirm that aggregates are correct.  

```{r}

# Check grand_total column 

crime_by_county |>
  group_by(county, year, grand_total)|> 
  summarise(check_grand_total = (murder+rape+robbery+agg_assault+breaking_and_entering+larceny_theft+motor_vehicle_theft))

# Check breaking_and_entering_per_100_000_people

crime_by_county |>
  group_by(county, year, breaking_and_entering_per_100_000_people)|> 
  summarise(check_breaking_and_entering_per_100_000_people = (breaking_and_entering/population)*100000)

# Check property_crime_percent_change using pct_change=((new-old)/old)*100. Phoned a friend on this one to figure out how to use lag(), your code below Derek. 

crime_by_county |> 
  group_by(county) |> 
  arrange(county, year, property_crime_percent_change) |> 
  mutate(check_property_crime_percent_change = (property_crime_totals - lag(property_crime_totals)) / lag(property_crime_totals) * 100) |> 
  filter(!is.na(check_property_crime_percent_change)) |> 
  select(county, year, property_crime_percent_change, check_property_crime_percent_change )

# Looks like the aggregates in the dataframe are accurate and can be used moving forward. 

```

### I am looking for trends across the crime categories over time. What types of crime have increased and what has decreased? What has stayed steady? Are there any outlying years that may have an interesting story behind them? Based on these trends, what crimes should Marylanders be most concerned about?

**Question** Let's start macro at the state level. How has crime in Maryland risen or fallen in the major categories since 1975? 

**Process Note** I got stuck in pretty short order because I didn't know how to visualize multiple variables in a line graph. I tried adding a second y= command but that didn't work. I figured that there's probably a function I don't know that will help me here so I asked ChatGPT: 

"I am working in R and using ggplot to visualize a dataset called trends_in_crime. The columns are year, total_rape,  total_murder, total_robbery,  total_agg_assault, total_breaking_and_entering, total_larceny_theft and total_motor_vehicle_theft.  I want to create a line graph where the x axis is year, the y axis is the count, and there is a line representing each of these columns: total_rape,  total_murder, total_robbery,  total_agg_assault, total_breaking_and_entering,  total_larceny_theft and total_motor_vehicle_theft." 

It told me, "# Assuming your data frame is named 'trends_in_crime'

# Reshape the data into long format
trends_long <- trends_in_crime %>%
  gather(key = "crime_type", value = "count", -year)

# Create a line graph using ggplot2
ggplot(trends_long, aes(x = year, y = count, color = crime_type)) +
  geom_line() +
  labs(title = "Crime Trends Over the Years",
       x = "Year",
       y = "Count",
       color = "Crime Type") +
  theme_minimal()" 
  
So here I learned that the function I needed was gather(key = "crime_type", value = "count", -year). I tweaked ChatGPT's code to meet my needs and that worked. Now I had a dataframe with a column called crime_type that included the types of crimes I wanted to include and a line graph with corresponding lines for those types. After looking at the graph, I went back and wrote a title representative of those findings. I also changed the category names to be capitalized for the sake of the graph aesthetics. 

**Answer** In general, crime in Maryland has trended down for nearly two decades. Most crime types appear to peak around 1980 and again around 1995. As of 2020 (the most recent year in the data), crime appears to be at record lows in almost every category. Rape and murder are consistently the most rare, but in this format it is difficult to tell whether rape and murder are at their lowest or if they're at similar levels to 1975. This probably deserves a closer look. 

Larceny theft continues to be the most prevalent crime, but it has also fallen the most dramatically. There could be an interesting story here: have policing practices around larceny theft changed? Has the definition of this crime changed? Or what factors have lead to a drastic decline in larceny theft over the past two years?  

```{r}

# Create new data frame summarizing the 7 crime categories. 

trends_in_crime <- crime_by_county |> 
  group_by(year)|> 
  summarise(
    Rape = sum(rape), 
    Murder = sum(murder), 
    Robbery = sum(robbery), 
    "Aggravated Assault" = sum(agg_assault), 
    "Breaking and Entering" = sum(breaking_and_entering), 
    "Larceny Theft" = sum(larceny_theft), 
    "Motor Vehicle Theft" = sum(motor_vehicle_theft)
  )

trends_in_crime

# Create line graph showing these trends over time. 

trends_in_crime_by_type <- trends_in_crime |>
  gather(key = "crime_type", value = "count", -year)

trends_in_crime_by_type |> 
  ggplot() +
  geom_line(aes(x = year, y = count, color = crime_type)) +
  labs(title = "Maryland Crime Dips to Record Lows",
       x = "Year",
       y = "Count",
       color = "Crime Type") +
  theme_minimal()

trends_in_crime_by_type

```

### Now let's compare violent crime and property crime statewide. 

**Question** How do trends in violent crime compare to property crime in Maryland? 

**Process Note** I decided to use Datawrapper for this question just to switch things up. 

**Answer** Datawrapper chart here: https://www.datawrapper.de/_/KSjKd/. Both violent and property crimes in Maryland fell to record lows by 2020. Property crime was cut in half and violent crime is down 15% from 1975 levels. As we know from the chart above, the major changes are in larceny theft and breaking and entering. 

```{r}

# Create new data frame summarizing violent crime and property crime. 

violent_versus_property <- crime_by_county |> 
  group_by(year)|> 
  summarise(
    "Violent Crime" = sum(violent_crime_total), 
    "Property Crime" = sum(property_crime_totals))

#Create a second graph showing violent crime versus property crime. Let's write to CSV and take it to DataWrapper this time. 

write_csv(violent_versus_property, "data/violent_versus_property.csv")

```

## Let's look specifically at the violent crime categories. 

**Question** What trends can we observe over time within in the violent crime category? 

**Answer** There are a couple of interesting things happening here. Aggravated assault and robbery both peaked sharply around 1993-2003 and have since fallen. Why would that be? It looks like rape and murder have maintained relatively stable over the past 4 decades, but again it's hard to really tell from this view. Upon closer look, 2020 had the 6th most murders of the timeframe, following a slew of peak years from 1992-1996. I think there is definitely a story about why murder jumped during a year where we were supposedly all supposed to be inside in our houses. Rape was relatively low in 2020, ranking 35 out of the past 46 years, so I suppose that may indicate a trend in the right direction. 

```{r}
# Sum up the 4 types of violent crime statewide. 

violent_crime_by_type <- crime_by_county |> 
  group_by(year)|> 
   summarise(
    Rape = sum(rape), 
    Murder = sum(murder), 
    Robbery = sum(robbery), 
    "Aggravated Assault" = sum(agg_assault)
  )

# Create line graph showing these trends over time. 

violent_crime_by_type_timeline <- violent_crime_by_type |>
  gather(key = "crime_type", value = "count", -year)
  
violent_crime_by_type_timeline|> 
  ggplot() +
  geom_line(aes(x = year, y = count, color = crime_type)) +
  labs(title = "Aggravated assault, robbery rates plummet in past 2 decades",
       x = "Year",
       y = "Count",
       color = "Crime Type") +
  theme_minimal()

# Look closer at murder. 

crime_by_county |> 
  group_by(year)|> 
  summarise(Murder = sum(murder)) |> 
  arrange(desc(Murder))

# Look closer at rape. 

crime_by_county |> 
  group_by(year)|> 
  summarise(Rape = sum(rape)) |> 
  arrange(Rape)

```

## Let's do the same analysis for property theft categories. 

**Question** What trends can we observe over time within in the property crime category? 

**Answer** As observed above, larceny has fallen the most drastically since mid-90's peaks. Motor vehicle theft appears to have jumped during the early 2000's as well. Breaking and entering has steadily declined. 

```{r}

# Sum up the 3 types of property crime statewide. 

property_crime_by_type <- crime_by_county |> 
  group_by(year)|> 
  summarise(
    "Breaking and Entering" = sum(breaking_and_entering), 
    "Larceny Theft" = sum(larceny_theft), 
    "Motor Vehicle Theft" = sum(motor_vehicle_theft)
  )

property_crime_by_type

# Create line graph showing these trends over time. 

property_crime_by_type_timeline <- property_crime_by_type |>
  gather(key = "crime_type", value = "count", -year)

property_crime_by_type_timeline |> 
  ggplot() +
  geom_line(aes(x = year, y = count, color = crime_type)) +
  labs(title = "Larceny down nearly 60% since 1995 peak",
       x = "Year",
       y = "Count",
       color = "Crime Type") +
  theme_minimal()

```

### Now let's look at just Prince George's County, assuming we are reporting for the sake of UMD students in College Park. 

**Question** Is crime in PG County trending similarly to the state as a whole? We have been hearing a lot about motor vehicle theft lately, can we see anything in particular from this dataset? 

**Answer** Crime in PG County is following statewide trends, including in motor vehicle theft. We would need to bring in more recent years to see if the tide is actually turning on motor vehicle theft or if the trend there is being exaggerated. 

```{r}

#Create a second graph showing violent crime versus property crime in PG County. 

trends_pg <- crime_by_county |> 
  filter(county == "PRINCE GEORGE'S COUNTY") |>
  group_by(year)|> 
  summarise(
    Rape = sum(rape), 
    Murder = sum(murder), 
    Robbery = sum(robbery), 
    "Aggravated Assault" = sum(agg_assault), 
    "Breaking and Entering" = sum(breaking_and_entering), 
    "Larceny Theft" = sum(larceny_theft), 
    "Motor Vehicle Theft" = sum(motor_vehicle_theft)
  )

trends_pg

# Create line graph showing these trends over time. 

trends_pg_timeline <- trends_pg |>
  gather(key = "crime_type", value = "count", -year)

trends_pg_timeline|> 
  ggplot() +
  geom_line(aes(x = year, y = count, color = crime_type)) +
  labs(title = "Crime in PG County plummets amidst statewide progress",
       x = "Year",
       y = "Count",
       color = "Crime Type") +
  theme_minimal()

```

```{r}

# It looks like something interesting is happening there with PG Count's motor vehicle theft rates being higher than the state as a whole. Let's look at that closer by merging the the statewide & PG County numbers. 

mv_statewide <- trends_in_crime |> 
  select(year, "Motor Vehicle Theft")|> 
  rename(Statewide = "Motor Vehicle Theft")

mv_pg <- trends_pg |> 
  select(year, "Motor Vehicle Theft")|> 
  rename("PG County" = "Motor Vehicle Theft")

mv_trends <- mv_statewide |> 
  left_join(mv_pg, join_by(year))

# Plot it. 

mv_trends_plot <- mv_trends |>
  gather(key = "Region", value = count, -year)

mv_trends_plot|> 
  ggplot() +
  geom_line(aes(x = year, y = count, color = Region)) +
  labs(title = "Motor vehicle theft in PG County mirrors statewide trend",
       x = "Year",
       y = "Count",
       color = "Crime Type", 
       caption = "source: Maryland State Police" ) +
  theme_minimal()

# Hmmm I guess not, looks like PG has been on trend with Maryland overall in this category. We seem to hear so much about it in the news lately, maybe it's gone back up in the past 3 years that aren't included in this dataset. 

```

### What could be causing property theft to go down so much? Is there anything we can compare from the ACS? 

**Question** Is there a relationship between the drop in property crime and the number of adults in the workforce? 

**Process Note** I had to use ChatGPT to create a curve in my final scatterplot. I asked it, "I am working in R with GGplot, my dataframe is called counties_census. I am visualizing the relationship between property crime rates and percent employment through a scatterplot. Here is my code: 

counties_census|>
  ggplot() + 
  geom_point(aes(y=property_crime_rate_per_100_000_people, x=pct_employed))+ 
  geom_text(aes(y=property_crime_rate_per_100_000_people, x=pct_employed, label=county), check_overlap = TRUE)+
  labs(
    title="Title TBD",
    y = "Property Crime",
    x = "Employment Rate",
    caption = "source: Maryland State Police")


Now I want to add a curve to this scatterplot to show the overall trend. Write me that code." 

It responded, "Sure, to add a curve to represent the overall trend between property crime rates and percent employment in your scatterplot, you can use a function like geom_smooth() in ggplot2. Here's the modified code:

counties_census |>
  ggplot() + 
  geom_point(aes(y = property_crime_rate_per_100_000_people, x = pct_employed)) + 
  geom_text(aes(y = property_crime_rate_per_100_000_people, x = pct_employed, label = county), check_overlap = TRUE) +
  geom_smooth(aes(y = property_crime_rate_per_100_000_people, x = pct_employed), method = "auto", se = FALSE) +
  labs(
    title = "Title TBD",
    y = "Property Crime",
    x = "Employment Rate",
    caption = "source: Maryland State Police"
  )"

**Answer** Frankly, this one surprises me. I thought that as the employment rate increased, property theft would decrease. It strikes me that this does not appear to be the case in Maryland over the past 15 years. Calvert, Carroll, Kent and Garrett counties appear to buck the trend. I'm not 100% confident in how I approached this question and would seek a second opinion before publication. 

```{r}

# Load census data from 2020, 2010, 2000. Let's look at employment status? https://acsdatacommunity.prb.org/discussion-forum/f/forum/395/acs-profile-data-variables
# says that DP03_0002PE EMPLOYMENT STATUS!!Population 16 years and over!!In labor force

md_2020  <- get_acs(geography='county', variables=c(adult__workforce='DP03_0002PE'), state='MD', year=2020)

# Rename the estimate and moe columns. Clean the county names so we can merge this dataset with crime_by_county. Put all the counties in all caps. Get rid of the GEOID and variable columns

md_2020 <- md_2020|> 
  rename(pct_employed = estimate)|> 
  rename(moe_employment = moe)|> 
  rename(county = NAME)|> 
  mutate(county = str_remove(county, ", Maryland")) |>  
  mutate(county = str_to_upper(county))|> 
  select(-GEOID, -variable)

# We need to establish what decade this info is from for our merge with crime_by_county. 

md_2020 <- md_2020 |> 
 mutate(census_year = 2020)

# Now let's merge with crime_by_county, years 2016-2020. 

crime_by_county_2020 <- crime_by_county |> 
 filter(year >= 2016 & year <= 2020)

# Now let's join crime_by_county_2020 with md_2020 

crime_by_county_2020 <- crime_by_county_2020 |> 
  left_join(md_2020, join_by(county))

# Great. Now let's repeat that process with 2015. 

md_2015  <- get_acs(geography='county', variables=c(adult__workforce='DP03_0002PE'), state='MD', year=2015)

md_2015  <- md_2015 |> 
  rename(pct_employed = estimate)|> 
  rename(moe_employment = moe)|> 
  rename(county = NAME)|> 
  mutate(county = str_remove(county, ", Maryland"))|> 
  mutate(county = str_to_upper(county))|> 
  mutate(census_year = 2015)|> 
   select(-GEOID, -variable)
 
crime_by_county_2015 <- crime_by_county |> 
 filter(year >= 2011 & year <= 2015)

crime_by_county_2015 <- crime_by_county_2015 |> 
  left_join(md_2015, join_by(county))

# And 2010. 

md_2010  <- get_acs(geography='county', variables=c(adult__workforce='DP03_0002PE'), state='MD', year=2010)

md_2010  <- md_2010 |> 
  rename(pct_employed = estimate)|> 
  rename(moe_employment = moe)|> 
  rename(county = NAME)|> 
  mutate(county = str_remove(county, ", Maryland"))|> 
  mutate(county = str_to_upper(county))|> 
  mutate(census_year = 2010)|> 
   select(-GEOID, -variable)

crime_by_county_2010 <- crime_by_county |> 
 filter(year >= 2006 & year <= 2010)

crime_by_county_2010 <- crime_by_county_2010 |> 
  left_join(md_2010, join_by(county))

# Combine. 

counties_census <- bind_rows(list(crime_by_county_2010, crime_by_county_2015, crime_by_county_2020))

# Let's look at the relationship between property crime and employment rates. 

counties_census|>
  ggplot() + 
  geom_point(aes(y=property_crime_rate_per_100_000_people, x=pct_employed))+ 
  geom_text(aes(y=property_crime_rate_per_100_000_people, x=pct_employed, label=county), check_overlap = TRUE)+
  geom_smooth(aes(y = property_crime_rate_per_100_000_people, x = pct_employed), method = "auto", se = FALSE) +
  labs(
    title="Property crime rises with MD employment rates, 2006-2020",
    y = "Property Crime",
    x = "Employment Rate",
    caption = "source: Maryland State Police")

```
### Pitch Memo: 

Overall, my data analysis paints an optimistic picture of crime trends in Maryland over the last 45 years. Crime has significantly dropped in most categories, particularly larceny and other types of property theft. Murder is the exception to this trend, as the 2020 murder rate was the highest outside of a peak of five years in the mid-90's. Prince George's County crime trends are in keeping with the rest of the state. As county-wide employment rates increase, property crime per capita actually appears to increase overall, but several counties defy that trend. Visualizations would be critical in supporting any analysis of crime trends, since the numbers in this dataset are so dense. 

Other questions I would like to explore in further analysis: 
* Do any of the trends change in 2021, 2022, 2023? 
* What do arrest rates look like in these categories? This dataset represents crimes, but does the Maryland Police track the number of arrests, or better yet, convictions? 
* How does Maryland compare to nation-wide trends? 



